---
title: "Data Structures"
teaching: 40
exercises: 15
questions:
- "How can I read data in R?"
- "How do I subset data to check it out?"
- "What are the basic data types in R?"
- "How do I represent categorical information in R?"
objectives:
- "To be aware of the different types of data."
- "To be able to subset using indices"
- "To begin exploring data frames, and understand how they are related to vectors, factors and lists."
- "To be able to ask questions from R about the type, class, and structure of an object."
keypoints:
- "Use `read.csv` to read tabular data in R."
- "The basic data types in R are double, integer, complex, logical, and character."
- "Use factors to represent categories in R."
source: Rmd
---
One of R's most powerful features is its ability to deal with tabular data -
such as you may already have in a spreadsheet or a CSV file. Let's start by
importing a dataset from your `data/` directory, called `stream-temps.csv`.

We can load this into R via the following:

```{r}
temps <- read.csv(file = "data/stream-temps.csv", stringsAsFactors=TRUE)
temps
```

The `read.table` function is used for reading in tabular data stored in a text
file where the columns of data are separated by punctuation characters such as
CSV files (csv = comma-separated values). Tabs and commas are the most common
punctuation characters used to separate or delimit data points in csv files. 
For convenience R provides 2 other versions of `read.table`. These are: `read.csv`
for files where the data are separated with commas and `read.delim` for files
where the data are separated with tabs. Of these three functions `read.csv` is
the most commonly used.  If needed it is possible to override the default 
delimiting punctuation marks for both `read.csv` and `read.delim`.


# Accessing elements using their indices

To extract elements of a vector we can give their corresponding index, starting from one. Here we are calling the first value from the vector `MonthlyMax` that is held within the `temps` dataframe.

```{r}
temps$MonthlyMax[1]
temps[1,4]
```

There are many ways to subset data, but these two options are the most common.  
We can select a column by using the `$` operator after the name of the dataframe. We then index into the column using `[]` to pull out a single value. Alternatively we can use `[]` with two values to select a value [*row*, *column*]. 

If we want to see multiple values we can show a *slice* of the vector using `:`

```{r}
temps$NoRWeST_ID[1:5]
temps$MonthlyMin[1:5]
```

Perhaps we want the data in Celcius instead of Fahrenheit, we can create a new column with the converted data:

```{r}
temps$MonthlyMin_C = (temps$MonthlyMin-32) * (5/9)
```

Or determine the departure from mean at a given location:
```{r}
temps$MonthlyMean[1] - temps$MonthlyMin[1]
```

But if you try combining a date with temperature what happens?

```{r}
temps$MonthlyMean[1] +  temps$MonthYear[1]
```

Understanding what happened here is key to successfully analyzing data in R.

## Data Types

AS you'd imagine, you cannot preform addition with a date, this concept of *data types* is important for programming, particularly when parsing environmental datasets. We can ask what type of data something is by:

```{r}
typeof(temps$MonthlyMean)
```

There are 5 main types: `double`, `integer`, `complex`, `logical` and `character`.

```{r}
typeof(3.14)
typeof(1L) # The L suffix forces the number to be an integer, since by default R uses float numbers
typeof(1+1i)
typeof(TRUE)
typeof('mayfly')
```

No matter how complicated our analyses become, all data in R is interpreted as one of these
basic data types. This strictness has some really important consequences.

Let's say someone has just started QAQC-ing the raw version of the data in the file `stream-temps-raw.csv`:

We'll load the data like we did before:

```{r}
temps_raw <- read.csv(file="data/stream-temps-raw.csv")
typeof(temps_raw$MonthlyMin)
```

Our temperatures aren't the double type anymore! If we try to do the same math
we did on them before, we run into trouble:

```{r}
temps_raw$MonthlyMean[1] - temps_raw$MonthlyMin[1]
```
What happened? When R reads a csv file into one of these tables, it insists that everything in a column be the same basic type; if it can’t understand *everything* in the column as a double, then *nothing* in the column can be a double. The table that R loaded our temperature data into is called a *data.frame*, and it is our first example of something called a data structure - that is, a structure which R knows how to build out of the basic data types.

We can see that it is a *data.frame* by calling the `class` function on it:

```{r}
class(temps)
```

In order to successfully use our data in R, we need to understand what the basic
data structures are, and how they behave. For now, we'll use the cleaned temperature data while we investigate this behavior further.

## Vectors and Type Coercion

To better understand this behavior, let's meet another of the data structures:
the *vector*.

```{r}
my_vector <- vector(length = 3)
my_vector
```

A vector in R is essentially an ordered list of things, with the special
condition that *everything in the vector must be the same basic data type*. If
you don't choose the datatype, it'll default to `logical`; or, you can declare
an empty vector of whatever type you like.


```{r}
another_vector <- vector(mode='character', length=3)
another_vector
```

You can check if something is a vector:

```{r}
str(another_vector)
```

The output from this command indicates a few things 1) the basic data type
found in this vector - in this case `chr`, character; 2) an indication of the
number of things in the vector - actually, the indexes of the vector, in this
case `[1:3]`; and 3) a few examples of what's actually in the vector - in this case
empty character strings. If we similarly do

```{r}
str(temps$SD)
```

we see that `temps$SD` is a vector, too - *the columns of data we load into R
data.frames are all vectors*, and that's the root of why R forces everything in
a column to be the same basic data type.

> ## Discussion 1
>
> Why is R so opinionated about what we put in our columns of data?
> How does this help us?
>
> > ## Discussion 1
> >
> > By keeping everything in a column the same, we allow ourselves to make simple
> > assumptions about our data; if you can interpret one entry in the column as a
> > number, then you can interpret *all* of them as numbers, so we don't have to
> > check every time. This consistency is what people mean when they talk about 
> > *clean data*; in the long run, strict consistency goes a long way to making 
> > our lives easier in R.
> {: .solution}
{: .discussion}

You can also make vectors with explicit contents with the combine function:

```{r}
combine_vector <- c(2,6,3)
combine_vector
```

Given what we've learned so far, what do you think the following will produce?

```{r}
quiz_vector <- c(2,6,'3')
```

This is something called *type coercion*, and it is the source of many surprises
and the reason why we need to be aware of the basic data types and how R will
interpret them. When R encounters a mix of types (here numeric and character) to
be combined into a single vector, it will force them all to be the same
type. Consider:

```{r}
coercion_vector <- c('a', TRUE)
coercion_vector
another_coercion_vector <- c(0, TRUE)
another_coercion_vector
```

The coercion rules go: `logical` -> `integer` -> `numeric` -> `complex` ->
`character`, where -> can be read as *are transformed into*. You can try to
force coercion against this flow using the `as.` functions:

```{r}
character_vector_example <- c('0','2','4')
character_vector_example
character_coerced_to_numeric <- as.numeric(character_vector_example)
character_coerced_to_numeric
numeric_coerced_to_logical <- as.logical(character_coerced_to_numeric)
numeric_coerced_to_logical
```

As you can see, some surprising things can happen when R forces one basic data
type into another! Nitty-gritty of type coercion aside, the point is: if your
data doesn't look like what you thought it was going to look like, type coercion
may well be to blame; make sure everything is the same type in your vectors and
your columns of data.frames, or you will get nasty surprises!

But coercion can also be very useful! For example, in our `temps_raw` data
`QAQC` is a numeric vector, but we know that the 1s and 0s actually represent
`TRUE` and `FALSE` (a common way of representing them). We should use the
`logical` datatype here, which has two states: `TRUE` or `FALSE`, which is
exactly what our data represents. We can 'coerce' this column to be `logical` by
using the `as.logical` function:

```{r}
temps_raw$QAQC[1:5]
temps_raw$QAQC <- as.logical(temps$QAQC)
temps_raw$QAQC[1:10]
```

The combine function, `c()`, will also append things to an existing vector:

```{r}
ab_vector <- c('a', 'b')
ab_vector
combine_example <- c(ab_vector, 'SWC')
combine_example
```

You can also make series of numbers:

```{r}
mySeries <- 1:10
mySeries
seq(10)
seq(1,10, by=0.1)
```

We can ask a few questions about vectors:

```{r}
sequence_ex <- seq(10)
head(sequence_ex, n=3)
tail(sequence_ex, n=4)
length(sequence_ex)
class(sequence_ex)
typeof(sequence_ex)
```

> ## Challenge 1
>
> What is an alternative way to show the beginning or end of a vector?
>
> > ## Challenge 1
> >
> > beginning: sequence_ex[1:3]
> > end: sequence_ex[7:10] or sequence_ex[7:length(sequence_ex)]
> > 
> > There are multiple ways to many things, this will come up often when you are looking at someone elses code, or examples online. It's a great way to find cool tricks and more efficient ways to do things!
> > 
> {: .solution}
{: .discussion}


Finally, you can give names to elements in your vector:

```{r}
my_example <- 5:8
names(my_example) <- c("a", "b", "c", "d")
my_example
names(my_example)
```

> ## Challenge 2
>
> Start by making a vector with the numbers 1 through 26.
> Multiply the vector by 2, and give the resulting vector
> names A through Z (hint: there is a built in vector called `LETTERS`)
>
> > ## Solution to Challenge 2
> >
> > ```{r}
> > x <- 1:26
> > x <- x * 2
> > names(x) <- LETTERS
> > ```
> {: .solution}
{: .challenge}

## Data Frames

We said that columns in data.frames were vectors:

```{r}
str(temps$MonthlyMax)
str(temps$QAQC)
```

These make sense. But what about

```{r}
str(temps$NoRWeST_ID)
```

## Factors

Another important data structure is called a *factor*. Factors usually look like
character data, but are typically used to represent categorical information. For
example, let's make a vector of strings labelling a few reaches of the Middle Snake:

```{r}
reaches <- c('MidSnake_1', 'MidSnake_2', 'MidSnake_1', 'MidSnake_3', 'MidSnake_2')
reaches
str(reaches)
```
We can turn a vector into a factor like so:

```{r}
reach_factors<- factor(reaches)
class(reach_factors)
str(reach_factors)
```

Now R has noticed that there are three possible categories in our data - but it
also did something surprising; instead of printing out the strings we gave it,
we got a bunch of numbers instead. R has replaced our human-readable categories
with numbered indices under the hood, this is necessary as many statistical
calculations utilise such numerical representations for categorical data:

```{r}
typeof(reaches)
typeof(reach_factors)
```
> ## Challenge 3
>
> Is there a factor in our `temps` data.frame? what is its name?
> Try using `?read.csv` to figure out how to keep text columns as character
> vectors instead of factors; then write a command or two to show that the factor
> in `temps` is actually a character vector when loaded in this way.
>
> > ## Solution to Challenge 2
> >
> > One solution is use the argument `stringAsFactors`:
> >
> > ```{r, eval = FALSE}
> > temps <- read.csv(file="data/stream-temps.csv", stringsAsFactors=FALSE)
> > str(temps$MonthYear)
> > ```
> >
> > Another solution is use the argument `colClasses`
> > that allow finer control.
> >
> > ```{r, eval = FALSE}
> > temps <- read.csv(file="data/stream-temps.csv", colClasses=c(NA, NA, "character"))
> > str(temps$MonthYear)
> > ```
> >
> > Note: help files can be difficult to understand; take your best guess based on semantic meaning,
> > even if you aren't sure, and remember google is a great resource.
> {: .solution}
{: .challenge}

## Lists - not as common in timeseries data sets ... but should prob include

## Matrices

Last but not least is the matrix. We can declare a matrix full of zeros:

```{r}
matrix_example <- matrix(0, ncol=6, nrow=3)
matrix_example
```

And similar to other data structures, we can ask things about our matrix:

```{r}
class(matrix_example)
typeof(matrix_example)
str(matrix_example)
dim(matrix_example)
nrow(matrix_example)
ncol(matrix_example)
```

> ## Challenge 4
>
> What do you think will be the result of
> `length(matrix_example)`?
> Try it.
> Were you right? Why / why not?
>
> > ## Solution to Challenge 4
> >
> > What do you think will be the result of
> > `length(matrix_example)`?
> >
> > ```{r}
> > matrix_example <- matrix(0, ncol=6, nrow=3)
> > length(matrix_example)
> > ```
> >
> > Because a matrix is a vector with added dimension attributes, `length`
> > gives you the total number of elements in the matrix.
> {: .solution}
{: .challenge}


> ## Challenge 5
>
> Make another matrix, this time containing the numbers 1:50,
> with 5 columns and 10 rows.
> Did the `matrix` function fill your matrix by column, or by
> row, as its default behaviour?
> See if you can figure out how to change this.
> (hint: read the documentation for `matrix`!)
>
> > ## Solution to Challenge 5
> >
> > Make another matrix, this time containing the numbers 1:50,
> > with 5 columns and 10 rows.
> > Did the `matrix` function fill your matrix by column, or by
> > row, as its default behaviour?
> > See if you can figure out how to change this.
> > (hint: read the documentation for `matrix`!)
> >
> > ```{r, eval=FALSE}
> > x <- matrix(1:50, ncol=5, nrow=10)
> > x <- matrix(1:50, ncol=5, nrow=10, byrow = TRUE) # to fill by row
> > ```
> {: .solution}
{: .challenge}