---
title: Data exploration in R
teaching: 35
exercises: 15
questions:
- "How can I use R to do standard statistics?"
objectives:
- "To be able to calculate means and standard deviations with subsets of data"
- "To be able to create a linear regression"
- "To be able to find packages that can support statistical analysis."
- "To be able to use ggplot2 to generate graphics."
keypoints:
- ""
- ""
- "Use `ggplot2` to create plots."

source: Rmd
---


```{r}
temps <- read.csv(file = "data/stream-temps.csv", stringsAsFactors=TRUE)
temps
```

One of the most common things we have to deal with when working with timeseries is converting dates into a format that is readable. The `as.Date` function converts from a character to R's `date` class, where the first argument is the character you want to convert, and the second is the structure of the date that is in the first argument. When I am trying something new that I dont have existing code to copy from, I often like to test things on a single value.

```{r}
temps$MonthYear[1]
```

We can see that the format is month - year, which according to https://www.r-bloggers.com/date-formats-in-r/ is equivilant to `%m-%Y`,. 

```{r}
as.Date(temps$date[1], "%m-%Y")
```

But for some reason we get `NA`, after searching around on the internet, I found that the `as.Date` function needs a day of the month. One way to get around this is to define the date as the first of the month using the `paste` function. 

```{r}
temps$date <- paste("01-",temps$MonthYear,sep="")
temps$date <- as.Date(temps$date, "%d-%m-%Y")
```

Now we can sort the data by month, year, or reach which will allow us to plot them more easily. 

```{r}
plot(temps$MonthYear[temps$NoRWeST_ID == 'MidSnake_174'], temps$MonthlyMean[temps$NoRWeST_ID == 'MidSnake_174'])
```


